\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-28 paper.tex
processed review remarks

\include{tikz}

@@

Supporting multiple intrusion detection (ID) algorithms imposes significant
overhead on Internet of Things (IoT) devices. Fusion of the underlying
algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem we introduce \NAME, a framework for the
development of efficient intrusion detection systems (IDS). It consists of a
domain specific language (DSL) that allows formally describing ID algorithms on
a functional level. A code generator then fuses these algorithms and produces
more optimally organized source code. This way, \NAME addresses the
heterogeneity of both the IoT and ID, as well as the limited resources of IoT
devices. A side-by-side comparison shows that generated code reduces memory
footprint, message passing overhead and execution time in comparison to a
straightforward sequential implementation of the same algorithms.

@@

The IoT presents a great potential to positively influence our daily work and
life: domotics, assisted living, e-health and enhanced learning are only a few
examples\cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes
\cite{aldrich2003smart} and personal data\cite{weber2010internet} to all these
interconnected devices, we also open them to everyone who is able to break the
virtual locks that protect them. IoT devices typically have limited batteries,
little processing power and memory is restricted to the bare minimum. These
properties make it hard to add any additional service, including security.

@@

In wireless networks of resource constraint devices, which make up an
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic\cite{zhang2000intrusion,
mishra2004intrusion}. If the outer security perimeter diminishes, every device
has to implement its own lines of defense. But making the IDS a local service
on each device requires local resources, which are the scarce commodity to IoT
devices.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms\cite{valero2012di} and offer the required basic
functional components to implement them\cite{krontiris2008lidea}. But they
don't offer a way to optimize the usage of resources. Nor do they support
systematic reuse to create different configurations for different devices.

% problem -> requirements for solution

The IoT is a heterogeneous environment, both in hardware and software.
Addressing this in a transparent and automated way is a key success factor for
any kind of development in this area. A solution that supports the integration
of multiple algorithms on IoT devices needs to consider the impact on their
resources. Different algorithms should be merged somehow to avoid accumulating
their impact.

% my approach merged into ...
% scientific contribution

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
identifies common data and functions, and organizes code in such a way that
redundant iterations, tests and computations are eliminated. Our results show
substantially improved memory usage, execution time and usage of the wireless
radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{positioning}
positions our approach as a development strategy for wireless devices, section
\ref{classification} takes a look at ID algorithms and classifies them. Section
\ref{pattern} looks for patterns in the implementation of these classes of
algorithms, details some of the ideas behind FCF and identifies the different
aspects of the problem space. Section \ref{design} describes the design we
applied to construct the framework, its domain specific language (DSL) and code
generator. Section \ref{evaluation} evaluates an implementation and application
of the framework. Section \ref{related} explores related work in the field of
DSLs for ID and code generation for the IoT. Finally, section \ref{conclusion}
summarizes our findings, draws conclusions and identifies topics for future
work.

\section{\NAME as a Development Strategy}
\label{positioning}

Figure \ref{fig:approach} shows a classification of several development
strategies that can be adopted for projects involving wireless devices.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\tdplotsetmaincoords{110}{-10}
\begin{tikzpicture}[scale=1,tdplot_main_coords]
  % axis
  \draw[very thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update cost$};
  \draw[very thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
  % points
  \coordinate (O)        at (0,0,0);
  \coordinate (ideal)    at (6,0,0);
  \coordinate (vm)       at (2,2,2);
  \coordinate (modular)  at (4,4,4);
  \coordinate (reconfig) at (4,2,4);
  \coordinate (full)     at (6,6,4);
  \coordinate (cg)       at (6,4,2.5);
  % "lines"
  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
  \draw[dashed] (vm)       -- (2,2,0);
  \draw[dashed] (modular)  -- (4,4,0);
  \draw[dashed] (full)     -- (6,6,0);
  \draw[dashed] (reconfig) -- (4,2,0);
  \draw[dashed] (cg)       -- (6,4,0);
  % "balls"
  \ball{6}{0}{0}{ideal}
  \ball{2}{2}{2}{virtual machine}
  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
  % 5. cg
  \tdplottransformmainscreen{6}{4}{2.5}
  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
    (\tdplotresx,\tdplotresy) circle (0.20)
    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
\end{tikzpicture}
}
\caption{Classification of Development Strategies for Wireless Devices (based on \cite{sadilek2007energy})}
\label{fig:approach}
\end{figure}

Three properties are used to classify the strategies\cite{sadilek2007energy}:
flexibility, update cost and development effort. The flexibility is a measure
for the number of applications that can be deployed on a device without
physical access. The update cost is the energy required to deploy a new version
of an application on a node. The development effort covers the time to write,
test and deploy an application. The \emph{ideal} situation is that were one
ends up with a maximal flexibility and no update nor development cost.

If we consider the trade off between flexibility and update cost, we can
distinguish two extremes: virtual machines and full native images. Virtual
machines allow for very small updates due to their concise byte code, but at
the same time are restricted by the limitations of their instruction set and
predefined capabilities. A full native image on the other hand has unlimited
flexibility, but requires a large transfer of the image over the network.

Two intermediate alternatives are possible: modular native image and
reconfigurable virtual machines. With modular native images, only modules are
changed, limiting the update size, but the framework to support it is static
and limits the possibilities. Reconfigurable virtual machines take this same
approach, but apply it to virtual machines, allowing to \emph{upgrade} the
instruction set of the virtual machine.

Choosing a development strategy, with this trade off take into account, will
also determine the development cost. Developing for a virtual machine will be
less demanding than targeting a native image.

With \NAME we opt to start from the full native image, allowing full
flexibility, but try to reduce both the update cost and the development cost.
Firstly, by applying code generation and functional code fusion, we optimize
the code organisation, which has a positive effect on the code size and hence
also on the update cost. Secondly, by enabling reuse on a functional level and
automating the writing phase, the development cost is also lowered.

@@

As in many other situations, a group is often stronger than the sum of its
individual components. Because not all devices are always actively
participating in the network, some might miss clues that would otherwise lead
them to detect intrusions. It is hardly impossible for a single device to
detect a complex attack by itself. Therefore a second important research topic
consists of cooperative algorithms to combine information from single
identities into a group-based decision about alleged
intrusions\cite{zhang2000intrusion}. We identify two ways by which this
cooperative decision making can be constructed and focus on the way
participants exchange information: broadcasting and interactive.

@@

  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
  to be implemented locally. This is illustrated by the reputation-based
  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
  the same principle of a watchdog also found in \cite{mishra2004intrusion}
  Using this algorithm, nodes can exchange information about the reputation of
  other nodes, using only broadcasted messages. Combining this information
  allows them to decide about their trust in a given node, based on more than
  their own, often partial, observations alone.

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system-based on the Guy Fawkes
  protocol\cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, a
  distributed decision can be made about the commonly identified intruder.

@@

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. It offers algorithms for exchanging
information about the software that is running on a device, and aims to
identify devices that can no longer prove that their content is unaltered. A
family of algorithms to implement this functionality are
SWATT\cite{seshadri2004swatt}, ICE/SCUBA\cite{seshadri2006scuba} and
SAKE\cite{seshadri2008sake}.

Software attestation is hard and many details can cause an algorithm to fail. A
discussion surrounding this topic, started by \cite{castelluccia2009difficulty}
in response to the previously mentioned papers, illustrates this. In
\cite{perrig2010refutation} the authors of the original papers counter many of
the objections made to their work, but the general feeling is that even in the
best conditions, there is always a way to circumvent the most ingenious
algorithm to perform software-based attestation.

Within the context of this paper, we will not consider software attestation as
a special class and exclude it from our scope, only focusing on detecting
intrusions and distributed decision making.

Figure \ref{fig:classification} visually summarizes the dual classification
presented above.

\begin{figure}[!ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  sibling distance=20mm,node distance=40mm
  ]
  \node (detection)             [align=center] {Detecting\\Intrusions}
    child {node (anomaly)       [align=center] {Anomaly}}
    child {node (signature)     [align=center] {Signature\\Misuse}}
    child {node (specification) [align=center] {Specification}};
  \node (decision)              [align=center,right=of detection] {Decision\\Making}
    child {node (broadcasting)  [align=center] {Broadcasting}}
    child {node (interactive)   [align=center] {Interactive}};
\end{tikzpicture}
}
  \caption{Classification of ID Algorithms}
  \label{fig:classification}
\end{figure}

@@

The latter is more complex: an algorithm based on signature or misuse detection
maybe sometimes can undeniably decide if an attack is real or not. But when
dealing with anomaly detection this line surely isn't that crisp. Especially in
distributed situations where even a neighboring device can not be trusted,
detecting intrusion is no absolute science. In this case one has to rely on
statistics and thresholds to decide when to drop trust in a given device.

Simply checking a threshold after updating some counter, is also not always a
good strategy. A simple example is that of a
watchdog\cite{mishra2004intrusion}, observing the actions of other devices.
There are valid reasons why a devices might miss the theoretical window in
which certain actions are expected. Simply reacting on such basic threshold
might result in false positives. A way to overcome this is to decouple the
detection of problems and the decision what to do about them.

@@

\begin{figure}[b]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

\begin{figure}[!ht]
\captionof{lstlisting}{Sequential implementation of ID algorithms\label{alg:id-algo-application}}
\begin{algorithmic}[1]
  \Let{$msg$}{$network$} \Comment{all observed messages}
  \ForEach{$algorithm \in algorithms$}
    \State \Call{algorithm.progress\_message}{msg}
  \EndFor
  \State \dots
  \Comment{at a given interval}
  \ForEach{$algorithm \in algorithms$}
    \State \Call{algorithm.do\_housekeeping}{}
  \EndFor
\end{algorithmic}
\end{figure}

@@

Given these patterns, we can now identify the problem areas that tend to result
in suboptimal code when combining multiple ID algorithms.

\subsection*{Loops and Common Data}

Two issues arise from the nested loops that originate from the combination of
the ID algorithm patterns and its application pattern. First, each algorithm,
by itself, will parse each observed message on the network, processing the same
byte stream, performing comparable operation to find similar patterns.

Second, there is common data: Each algorithm also has to store information
about nodes within its own scope, resulting in a lot of duplicated information
in memory.

Reimplementing the algorithms manually to eliminate these issues is not a valid
option. This development cost typically isn't a one-time cost, e.g. in case a
new version of the algorithm is released, or simply when a different set of
algorithms is selected due to a changing risk analysis. Finally, changing, or
simply implementing the algorithms from scratch, also holds the risk of making
mistakes.

\subsection*{Network Usage and Heterogeneity}

A third issue is hidden in the use of the network. The calls to the \ttt{SEND}
function are scattered around the algorithm, causing the wireless radio to be
accessed multiple times, keeping it active and energy consuming.

The heterogeneity of the IoT devices, results in many different software
stacks, each with their own network API. This also explains why there are very
few implementations readily available to developers: there is no common
platform to target for researchers.

Even using a unified software library would not only require a consistent
implementation of it, but would also limit the operational scope of the
research implementation, still restricting it to a single language.

@@

\subsection*{The Meaning of Loops}

From the design patterns we learned that loops are a major contributor to the
problem of algorithm fusibility. Loops in program code are technical constructs
and in fact often hide the actual intended functionality.

To illustrate this, we take the example of parsing network messages. Each
algorithm deals with parsing in the same way: loop over the bytes in the
payload and look for a pattern that identifies what the algorithm is looking
for. This is in fact a technical implementation choice. The underlying
functional goal is to react when such pattern is present in a message. The loop
has \emph{activated} this functionality, but the real goal is to
\emph{passively} react on an event, not \emph{actively} look for it.

This is how \NAME removes the need for loops: by introducing event handling,
constructions with loops are transformed to their functional meaning, allowing
much easier identification and fusion of common functionalities.

@@

\subsection*{A DSL for Nodes}

The second problem that could be observed from the patterns, was that of common
data. All algorithms deal with information about neighboring devices. To
centralize this information \NAME provides a unified \texttt{nodes} concept as
part of its DSL and exposes it as an object to the implementor.

Nodes can be extended with properties only available to the algorithm that
defines them. This way only one instance for every known devices will be
present in memory. Finally, nodes can define a scope for some function to be
executed in. This way, nodes also offer a way to iterate all known devices,
again without the need for a loop.

\subsection*{Unified Messaging}

@@

@@

\lstdefinelanguage{foo-lang}{
  emph={module,const,from,import,extend,with,after,do,function,@every,having},
  emphstyle={\textbf},
  morecomment=[l]{//},
  moredelim=[is][]{@@}{@@},       % temp solution to allow keywords without emph
  moredelim=[is][\emph]{!!}{!!}   % temp solution to hightlight #atoms
}

@@

\subsection*{Code Generator}

@@

A semantic model \cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations can now be applied to create a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit actual source code. Figure
\ref{fig:code-generation} shows the overall procedure.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  node distance=8mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                 {DSL};
  \node (sm)   [representation,right=of dsl,align=center] {Semantic\\Model};
  \node (cm)   [representation,right=of sm,align=center]  {Code\\Model};
  \node (code) [resource,right=of cm]        {Code};

@@

\subsection*{Semantic Model}

@@

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  edge from parent fork down,
  sibling distance=5mm, level distance=20mm, node distance=6mm
  ]
  \node (model) [class] {Model}
      child {node (module) [class] {Module} edge from parent [composite]
        child {node (strategy) [class,align=center,xshift=-9mm] {Execution\\Strategy} edge from parent [composite]
          child {node(every) [class,xshift=-9mm] {Every} edge from parent [subclass]}
          child {node (when) [class,xshift=9mm] {When} edge from parent [subclass]}
        }
        child {node (function) [class,xshift=9mm] {Function} edge from parent [composite]}
      };
  \node (constant)  [class,right=of module]   {Constant};
  \node (scope)     [class,left=of strategy]  {Scope};
  \node (statement) [class,right=of function] {Statement};
  \node (nodes)     [class,left=of module]    {Nodes\\Domain};

>>>>

\include{tikz}
\include{foo-lang}

@@

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns, anomalies,\dots that can indicate malicious
behavior. But supporting these intrusion detection algorithms imposes
significant overhead on resource-restricted devices such as those targeting the
Internet of Things. Fusion of the algorithms, to optimize the use of the shared
resources, can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem we introduce \NAME, a framework for the
development of efficient intrusion detection systems. It consists of a domain
specific language, to formally describing intrusion detection algorithms on a
functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that generated code reduces memory
footprint, message passing overhead and execution time in comparison to a
straightforward sequential implementation of the same algorithms.

@@

The internet of things (IoT) presents a great potential to positively influence
our daily work and life: domotics, assisted living, e-health and enhanced
learning are only a few examples\cite{atzori2010internet}. But with this
potential the IoT also presents a significant threat: by opening our smart
homes and personal data to all these interconnected devices, we also open them
to everyone who is able to break the virtual locks that protect them. IoT
devices typically have limited batteries, little processing power and memory is
restricted to the bare minimum. These properties make it hard to add any
additional service, including security.

@@

In wireless networks of resource constraint devices, which make up an
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic\cite{mishra2004intrusion}. If the
outer security perimeter diminishes, every device has to implement its own
lines of defense. But making the IDS a local service on each device requires
local resources, which are the scarce commodity to IoT devices.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms\cite{valero2012di} and offer the required basic
functional components to implement them\cite{krontiris2008lidea}. The first
aspect that is missing, is that they don't offer a way to optimize the usage of
resources. A solution that supports the integration of multiple algorithms on
IoT devices needs to consider the impact on their resources and avoid
accumulating their impact.

Secondly, they don't support systematic reuse to create different
configurations for different devices, both dealing with hard- and software.
Addressing this in a transparent and automated way is a key success factor for
any kind of development in this area.

% scientific contribution

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
identifies common data and functions, and organizes code in such a way that
redundant iterations, tests and computations are eliminated. Results show
substantially improved memory usage, execution time and usage of the wireless
radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
for patterns in the implementation of these classes of algorithms, details some
of the ideas behind FCF and identifies the different aspects of the problem
space. Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation and application of the framework.
Section \ref{related} explores related work in the field of DSLs for ID and
code generation for the IoT. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

@@

As in many other situations, a group is often stronger than the sum of its
individual components. Because not all devices are always actively
participating in the network, some might miss clues that would otherwise lead
them to detect intrusions. It is hardly impossible for a single device to
detect a complex attack by itself. Therefore a second important research topic
consists of cooperative algorithms to combine information from single
identities into a group-based decision about alleged intrusions. We identify
two ways by which this cooperative decision making can be constructed and focus
on the way participants exchange information: broadcasting and interactive.

@@

  \item[Broadcasting] can sometimes allow a distributed, cooperative algorithm
  to be implemented locally. This is illustrated by the reputation-based
  detection algorithm introduced in \cite{ganeriwal2008reputation}, based on
  the same principle of a watchdog also found in \cite{mishra2004intrusion}.
  Using this algorithm, nodes can exchange information about the reputation of
  other nodes, solely with broadcasted messages. Combining this information
  allows them to decide about their trust in a given node, based on more than
  their own, often partial, observations alone.

  \item[Interactive] communication allows algorithms to exchange information to
  reach a consensus. In \cite{krontiris2009cooperative}, the authors first
  present a theoretical foundation to analyze cooperative algorithms. Given
  these foundations they continue to present an algorithm consisting of 5
  phases. It essentially implements a voting system-based on the Guy Fawkes
  protocol\cite{anderson1998new} to identify an intruder. It enables the
  exchange of suspected intruder information and allows distributed
  authentication of those \emph{votes}. Based on these authenticated votes, a
  distributed decision can be made about the jointly identified intruder.

@@

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. It offers algorithms for exchanging
information about the software that is running on a device, and aims to
identify devices that can no longer prove that their content is unaltered. A
good example is SAKE\cite{seshadri2008sake}.

Software attestation is hard and many details can cause an algorithm to fail.
The general feeling is that even in the best conditions, there is always a way
to circumvent the most ingenious algorithm to perform software-based
attestation.

Within the context of this paper, we will not consider software attestation as
a special class and exclude it from our scope, only focusing on detecting
intrusions and distributed decision making. Figure \ref{fig:classification}
visually summarizes this dual classification.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  sibling distance=20mm,node distance=40mm
  ]
  \node (detection)             [align=center] {Detecting\\Intrusions}
    child {node (anomaly)       [align=center] {Anomaly}}
    child {node (signature)     [align=center] {Signature\\Misuse}}
    child {node (specification) [align=center] {Specification}};
  \node (decision)              [align=center,right=of detection] {Decision\\Making}
    child {node (broadcasting)  [align=center] {Broadcasting}}
    child {node (interactive)   [align=center] {Interactive}};
\end{tikzpicture}
}
  \caption{Classification of ID Algorithms}
  \label{fig:classification}
\end{figure}

@@

The latter is more complex: an algorithm based on signature or misuse detection
can maybe undeniably decide if an attack is real or not. But when dealing with
anomaly detection this line surely isn't that crisp. Especially in distributed
situations where even a neighboring device can not be trusted, detecting
intrusion is no absolute science. In this case one has to rely on statistics
and thresholds to decide when to drop trust in a given device.

But simply checking a threshold after updating some counter, is not a good
strategy. A simple example is that of a watchdog\cite{mishra2004intrusion},
observing the actions of other devices. There are valid reasons why a devices
might miss the theoretical window in which certain actions are expected. Simply
reacting on such basic threshold might result in false positives. A way to
overcome this is to decouple the detection of problems and the decision what to
do about them.

@@

\begin{figure}[ht]
\captionof{lstlisting}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyze byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\end{figure}

@@

\begin{figure}[ht]
\captionof{lstlisting}{Application pattern of multiple ID algorithms\label{alg:id-algo-application}}
\begin{algorithmic}[1]
  \Let{$msg$}{$network$} \Comment{all observed messages}
  \ForEach{$algorithm \in algorithms$}
    \State \Call{algorithm.progress\_message}{msg}
  \EndFor
  \State \dots
  \Comment{at a given interval}
  \ForEach{$algorithm \in algorithms$}
    \State \Call{algorithm.do\_housekeeping}{}
  \EndFor
\end{algorithmic}
\end{figure}

@@

\subsection*{Inherent Problems}
\label{pattern-problems}

Given these patterns, we can now identify the problem areas that tend to result
in suboptimal code when combining multiple ID algorithms. Two issues arise from
the nested loops that originate from the combination of the ID algorithm
pattern and the application pattern: loops and common data. A third issue is
the use of the network.

  \item[Loops] by themselves are not a problem, but when different loops
  perform the same operations in sequence, an overhead might be incurred. An
  example from the ID algorithm pattern is the handling of messages: each
  algorithm parses each observed message on the network, processing the same
  byte stream, performing comparable operations to find similar patterns.

  \item[Common data] is cross-cutting the different algorithms, that have to
  store information about nodes within its own scope. This results in a lot of
  duplicated information in memory that also needs to be maintained.

  \item[The network] allows algorithms to operate on a distributed scale. But
  calls to the \ttt{SEND} function are scattered around the algorithms, causing
  the wireless radio to be accessed multiple times, keeping it active and
  energy consuming more than strictly needed.

Reimplementing the algorithms manually to eliminate these issues is not a valid
option. This development cost typically isn't a one-time cost, e.g. in case a
new version of the algorithm is released, or simply when a different set of
algorithms is selected due to a changing risk analysis. Also, changing, or
simply implementing the algorithms from scratch, holds the risk of making
mistakes.

The heterogeneity of IoT devices results in many different software stacks,
each with e.g. their own network API. This also explains why there are very few
implementations readily available to developers: there is no common platform to
target for researchers. Even using a unified software library would not only
require a consistent implementation of it, but would also limit the operational
scope of the research implementation, still restricting it to a single language.

@@

Choosing for a DSL and code generator is choosing a development strategy. This
implies several choices with respect to flexibility, update cost and
development effort. We will first position \NAME and motivate its design with
respect to these trade offs (\ref{positioning}). Second, we will look at the
principal design components of the DSL and how it deals with loops and common
data (\ref{dsl-design}). Third, the structure of the code generator will be
discussed (\ref{code-generator-design}). Finally, fourth, the supporting
software library is introduced (\ref{software-lib-design}).

\subsection{\NAME as a Development Strategy}
\label{positioning}

Figure \ref{fig:positioning} shows a classification of several development
strategies that can be adopted for projects involving wireless devices.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\tdplotsetmaincoords{110}{-10}
\begin{tikzpicture}[scale=1,tdplot_main_coords]
  % axis
  \draw[very thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
  \draw[very thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
  \draw[very thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
  % points
  \coordinate (O)        at (0,0,0);
  \coordinate (ideal)    at (6,0,0);
  \coordinate (vm)       at (2,2,2);
  \coordinate (modular)  at (4,4,4);
  \coordinate (reconfig) at (4,2,4);
  \coordinate (full)     at (6,6,4);
  \coordinate (cg)       at (6,4,2.5);
  % "lines"
  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
  \draw[dashed] (vm)       -- (2,2,0);
  \draw[dashed] (modular)  -- (4,4,0);
  \draw[dashed] (full)     -- (6,6,0);
  \draw[dashed] (reconfig) -- (4,2,0);
  \draw[dashed] (cg)       -- (6,4,0);
  % "balls"
  \ball{6}{0}{0}{ideal}
  \ball{2}{2}{2}{virtual machine}
  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
  % 5. cg
  \tdplottransformmainscreen{6}{4}{2.5}
  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
    (\tdplotresx,\tdplotresy) circle (0.20)
    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
  % "box around our solution" ;-)
  \tdplottransformmainscreen{0}{0}{0}
  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);

\end{tikzpicture}
}
\caption{Classification of Development Strategies for Wireless Devices (based on \cite{sadilek2007energy})}
\label{fig:positioning}
\end{figure}

Three properties are used to classify the strategies\cite{sadilek2007energy}:
flexibility, update cost and development effort.

  \item[Flexibility] is a measure for the number of applications that can be
  deployed on a device without physical access.

  \item[Update cost] is equivalent to the energy required to deploy a new
  version of an application on a device.

  \item[Development effort] covers the time to write, test and deploy an
  application.

The \emph{ideal} situation is that were one ends up with a maximal flexibility
and no update nor development cost.

If we consider the trade off between flexibility and update cost, we can
distinguish two extremes: \emph{virtual machines} and \emph{full native
images}. Virtual machines allow for very small updates due to their concise
byte code, but at the same time are restricted by the limitations of their
instruction set and predefined capabilities. A full native image on the other
hand has unlimited flexibility, but requires transfer of a larger image over
the network.

Two intermediate alternatives are possible: \emph{modular native image} and
\emph{reconfigurable virtual machines}. With modular native images, only
modules are changed, limiting the update size, but the framework to support it
is static and limits the possibilities. Reconfigurable virtual machines take
this same approach, but apply it to virtual machines, allowing to
\emph{upgrade} the instruction set of the virtual machine.

Choosing a development strategy also determines the development cost: e.g.
developing for a virtual machine will be less demanding than targeting a native
image.

With \NAME we opt to start from the full native image, allowing full
flexibility, but try to reduce both the update cost and the development cost.
Firstly, by applying code generation and functional code fusion, we optimize
the code organization, which has a positive effect on the code size and hence
also on the update cost. Secondly, by enabling reuse on a functional level and
automating the writing phase, the development cost is also lowered.

\subsection{Domain Specific Language}
\label{dsl-design}

Using a DSL, and especially an external DSL, as a foundation for any solution
requires special care. It's important to remember that a DSL is merely a
\emph{thin veneer over a (semantic) model}\cite{fowler2010domain}. Its sole
purpose should be the easing of filling out the model. In the case of \NAME
this easing focuses on avoiding constructions that result in problems as
suggested in \ref{pattern-problems} concerning loops, common data and the use
of the network.

\subsubsection*{Loops}

Loops are a major contributor to the problem of algorithm fusibility. Loops in
program code are technical constructs and in fact often hide the actual
intended functionality.

To illustrate this, we take the example of parsing network messages. Each
algorithm deals with parsing in the same way: loop over the bytes in the
payload and look for a pattern that identifies what the algorithm is looking
for. This is in fact a technical implementation choice. The underlying
functional goal is to react when such pattern is present in a message. The loop
has \emph{activated} this functionality, but the real goal is to
\emph{passively} react on an event, not \emph{actively} look for it. Using
event handling, constructions with loops are transformed to their functional
meaning, allowing identification and fusion of common functionalities.

@@

\subsubsection*{The Nodes Domain}

All algorithms deal with information about neighboring devices. To centralize
this information \NAME provides a unified \ttt{nodes} concept as part of its
DSL and exposes it as an object to the implementor.

Nodes can be extended with properties only available to the algorithm that
defines them. This way only one instance for every known device will be present
in memory. Finally, nodes can define a scope for some function to be executed
in. This way, nodes also offer a way to iterate all known devices, again
without the need for a loop.

\subsubsection*{Unified Messaging}
\label{dsl-unified-msg}

Access to the network that is scattered around an algorithm, causes the
wireless radio to be more activated than needed. Delaying the sending of
messages and grouping them, can lower this overhead.

@@

\subsubsection*{An Example}

@@

@@

\subsection{Code Generator}
\label{code-generator-design}

@@

A semantic model\cite{fowler2010domain}, tied closely to the DSL itself, can
fulfill this role. Semantic transformations can now be applied to create a
code-oriented model, comparable to the idea of CodeDom. From there on,
syntactic transformations an emit actual source code. Figure
\ref{fig:code-generation} shows the overall procedure.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  node distance=7mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                 {DSL};
  \node (sm)   [representation,right=of dsl,align=center] {Semantic\\Model};
  \node (cm)   [representation,right=of sm,align=center]  {Code\\Model};
  \node (code) [resource,right=of cm]        {Code};

@@

\subsubsection*{Semantic Model}

@@

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  edge from parent fork down,
  sibling distance=5mm, level distance=20mm, node distance=6mm
  ]
  \node (model) [class,fill=black!10!white] {Model}
      child {node (module) [class,fill=black!10!white] {Module} edge from parent [composite]
        child {node (strategy) [class,align=center,xshift=-10.5mm,fill=black!10!white] {Execution\\Strategy} edge from parent [composite]
          child {node(every) [class,xshift=-9mm,fill=black!10!white] {Every} edge from parent [subclass]}
          child {node (when) [class,xshift=9mm,fill=black!10!white] {When} edge from parent [subclass]}
        }
        child {node (function) [class,xshift=10.5mm] {Function} edge from parent [composite]}
      };
  \node (constant)  [class,right=of module]   {Constant};
  \node (scope)     [class,left=of strategy]  {Scope};
  \node (statement) [class,right=of function] {Statement};
  \node (nodes)     [class,left=of module,xshift=-13mm]    {Nodes\\Domain};

  \draw [aggregate]  (module)   -- (nodes);
  \draw [composite]  (module)   -- (constant);
  \draw [dependency] (strategy) -- (scope);
  \draw [composite]  (function) -- (statement);
